---
title: 'Statistical Inference Course Project - Part 2: Basic Inferential Data Analysis'
author: "Ana Cortazzo"
date: "`r Sys.Date()`"
output:
  html_document: null
  pdf_document: default
---

# Synopsis

In this work, the data of tooth growth in Guinea pigs, belonging to the dataset of R, are studied. First, an explotarotio analysis of the data is carried out, followed by a statistical inference analysis to detect correlation between the variables of the study. 

# Data Processing

## Loading data

```{r message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(datasets)
data(ToothGrowth)
attach(ToothGrowth)
```

This data show the effect of vitamin C on tooth growth in Guinea Pigs. The response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, orange juice (coded as OJ) or ascorbic acid (a form of vitamin C and coded as VC).

A data frame with 60 observations on 3 variables.

- `len`	numeric	Tooth length
- `supp`	factor	Supplement type (VC or OJ).
- `dose`	numeric	Dose in milligrams/day


## Analyzing data

First look at the basic structure of data and a small summary statistic of the variable `len`.

```{r message=FALSE, warning=FALSE}
str(ToothGrowth)
summary(len)
```


```{r message=FALSE, warning=FALSE}
ggplot(ToothGrowth, aes(dose, len)) +
    geom_point(color = "red") + 
    facet_grid(.~supp) +
    geom_smooth(method = "lm") +
    labs(title = "Tooth length vs. dose", x = "Dose (mg/day)", y = "Tooth length")
```

It is of interest to know how the `len` variable behaves when we separate by `supp` or `dose`. It can be seen that there is a large variation in relation to the mean for the different groups observed. 

```{r message=FALSE, warning=FALSE}
ToothGrowth %>% group_by(supp) %>% summarise(n=n(), mean=mean(len), sd=sd(len))
ToothGrowth %>% group_by(dose) %>% summarise(n=n(), mean=mean(len), sd=sd(len))
ToothGrowth %>% group_by(supp,dose) %>% summarise(n=n(), mean=mean(len), sd=sd(len))
```



# Inferential Data Analysis


To carry out the statistical analysis, we used the R `t.test`, since it allows us to analyze the confidence intervals and test hypotheses. First we compared the tooth growth according to the supplement used, and then according to the dose.

## `t.test` to compare tooth growth by supplement type

To perform the test the following assumptions are made: 

- the groups have different variances

- the groups are independent, they are not treated as paired

- null hypothesis ($H_0$): there is no difference in `len` by `dose`

- $\alpha = 0,05$


```{r message=FALSE, warning=FALSE}
len_by_supp <- t.test(len~supp, data = ToothGrowth, paired = FALSE, alternative = "two.sided", var.equal = FALSE, conf.level = .95)
len_by_supp
```

The confidence interval is `r len_by_supp$conf.int` with a $95\%$ probability, it also contains $0$, which implies that it is possible that both means are equal. The null hypothesis cannot be rejected since p-value (=`r len_by_supp$p.value`) > $\alpha$, therefore it cannot be affirmed if there is a difference in tooth growth depending on the supplement used.


## `t.test` to compare tooth growth by dose

The comparison is made in three groups:

- **Case 1**: dose = 0.5 mg/day with dose = 1 mg/day

- **Case 2**: dose = 0.5 mg/day with dose = 2 mg/day

- **Case 3**: dose = 1 mg/day with dose = 2 mg/day

The same assumptions are made as before.

### Case 1

Select the required data
```{r message=FALSE, warning=FALSE}
dose_0.5_1 <- ToothGrowth  %>% filter(dose != 2)
```

```{r message=FALSE, warning=FALSE}
len_by_dose1 <- t.test(len~dose, data = dose_0.5_1, paired = FALSE, alternative = "two.sided", var.equal = FALSE, conf.level = .95)
len_by_dose1
```
The confidence interval is `r len_by_dose1$conf.int` with a $95\%$ probability. 

p-value (=`r len_by_dose1$p.value`) is significantly less than $\alpha$, the null hypothesis can be rejected, which can be understood as meaning that there is a difference in the average growth of tooth when a dose of 0.5 mg/day and a dose of 1 mg/day are applied.

### Case 2

Select the required data
```{r message=FALSE, warning=FALSE}
dose_0.5_2 <- ToothGrowth  %>% filter(dose != 1)
```

```{r message=FALSE, warning=FALSE}
len_by_dose2 <- t.test(len~dose, data = dose_0.5_2, paired = FALSE, alternative = "two.sided", var.equal = FALSE, conf.level = .95)
len_by_dose2
```
The confidence interval is `r len_by_dose2$conf.int` with a $95\%$ probability. 

p-value (=`r len_by_dose2$p.value`) is significantly less than $\alpha$, the null hypothesis can be rejected, which can be understood as meaning that there is a difference in the average growth of tooth when a dose of 0.5 mg/day and a dose of 2 mg/day are applied.

### Case 3

Select the required data
```{r message=FALSE, warning=FALSE}
dose_1_2 <- ToothGrowth  %>% filter(dose != 0.5)
```

```{r message=FALSE, warning=FALSE}
len_by_dose3 <- t.test(len~dose, data = dose_1_2, paired = FALSE, alternative = "two.sided", var.equal = FALSE, conf.level = .95)
len_by_dose3
```
The confidence interval is `r len_by_dose3$conf.int` with a $95\%$ probability. 

p-value (=`r len_by_dose3$p.value`) is significantly less than $\alpha$, the null hypothesis can be rejected, which can be understood as meaning that there is a difference in the average growth of tooth when a dose of 1 mg/day and a dose of 2 mg/day are applied.

# Conclusions

After performing a small statistical inference analysis, it was possible to observe that there is a significant correlation between increasing the dose of vitamin C and the growth of tooth growth in Guinea pigs. This could be clearly observed when the three doses were compared by hypothesis testing. However, the relationship of tooth growth with the method of administering this vitamin is not clear; the analysis performed does not allow identifying significant differences.